---
title: "first script"
author: "Annie Freeman"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load the tools
library(tidyverse)
library(tidycensus)

# Set your Census key (it pulls it from the secret file we made)
census_api_key(Sys.getenv("CENSUS_API_KEY"))

# A quick test: Pull median age for all states in 2022
test_data <- get_acs(
  geography = "state",
  variables = "B01002_001", 
  year = 2022
)

# Look at your data
head(test_data)
```


Pulling the 2010 vs 2022 Data
For your project, we need two "snapshots" to compare. We will pull the Median Earnings for the 25–34 age group (this is the closest standard bracket to your 22–34 target).

```{r}
# This creates a searchable list of all 2022 ACS variables
v22 <- load_variables(2022, "acs5", cache = TRUE)

# View it to search (a new tab will open)
View(v22)
```


Inflation 
A dollar in 2010 bought a lot more than a dollar does in 2026. To answer if workers are "worse off," we have to adjust the 2010 wages.This is where the FRED API comes in. In our next script, we will pull the Consumer Price Index (CPI) to calculate the "Real" value of those 2010 wages using this formula:$$\text{Real Wage}_{2010} = \text{Nominal Wage}_{2010} \times \left( \frac{\text{CPI}_{2026}}{\text{CPI}_{2010}} \right)$$



Finding the "Age" Variables
In the Census, "Median Earnings" by specific age brackets isn't always in table B20017. The better table for this project is likely B19049 (Median Household Income by Age of Householder) or searching for individual earnings specifically for those age groups.

```{r}
# This creates a searchable list of all 2022 ACS variables
v22 <- load_variables(2022, "acs5", cache = TRUE)

# View it to search (a new tab will open)
View(v22)
```

Pull Code
Now that we know the table, let’s pull the data for those two age brackets. This code pulls both the "Under 25" and "25-44" groups so you can compare them.

```{r}
# Define the variables for the two age groups
target_vars <- c(
  "Under_25" = "B19049_002",
  "Age_25_44" = "B19049_003"
)

# Pull 2010 Data
income_2010 <- get_acs(
  geography = "state",
  variables = target_vars,
  year = 2010,
  survey = "acs5"
) %>% mutate(report_year = 2010)

# Pull 2022 Data (Latest available)
income_2022 <- get_acs(
  geography = "state",
  variables = target_vars,
  year = 2022,
  survey = "acs5"
) %>% mutate(report_year = 2022)

# Merge them together
full_income_comparison <- bind_rows(income_2010, income_2022)
```
Open that final_income_data table. Compare a state’s 2010 real_income to its 2022 estimate.
- If the 2010 Real Income is HIGHER than the 2022 Income: It means workers back then were effectively "richer" because their money went further.
- If the 2022 Income is HIGHER: It means wages have actually managed to outrun the cost of groceries and gas.




Right now, you’re looking at a big spreadsheet that answers the question: "How much money did young people actually take home in 2010 vs. 2022?"


It's going to take those 2010 numbers and say: "If this Alabama worker wanted to live the same lifestyle in 2022, they would have needed $X amount.



# FRED Inflation Data

```{r}
library(fredr)

# 1. Set your FRED key (from your secret file)
fredr_set_key(Sys.getenv("FRED_API_KEY"))

# 2. Pull CPI data for 2010 and 2022
# 'CPIAUCSL' is the standard code for the Consumer Price Index
cpi_data <- fredr(
  series_id = "CPIAUCSL",
  observation_start = as.Date("2010-01-01"),
  observation_end = as.Date("2022-12-31"),
  frequency = "a" # "a" means Annual average
)

# 3. Simplify the table to just the Year and the Index Value
cpi_clean <- cpi_data %>%
  mutate(report_year = year(date)) %>%
  select(report_year, cpi_value = value)

# Look at your CPI table
print(cpi_clean)
```


```{r}
# A very simple test call
fredr(series_id = "GNP")
```

```{r}
nchar(Sys.getenv("FRED_API_KEY"))
```

```{r}
# Pulling CPI data with "safer" formatting
cpi_data <- fredr(
  series_id = "CPIAUCSL",
  observation_start = as.Date("2010-01-01"),
  observation_end = as.Date("2022-12-31")
)

# If that works, we can "clean" it afterwards
cpi_clean <- cpi_data %>%
  mutate(report_year = year(date)) %>%
  group_by(report_year) %>%
  summarize(cpi_value = mean(value)) # This averages the monthly data into a yearly number
```



Now we move from "gathering ingredients" to "cooking the meal." We are going to take that inflation data and use it to adjust your 2010 Census numbers.

#The "Real Income" Calculation
This code will join our income data with your inflation data and do the math.


```{r}
# 1. Get the specific CPI values for 2010 and 2022 to use in our formula
cpi_2010 <- cpi_clean$cpi_value[cpi_clean$report_year == 2010]
cpi_2022 <- cpi_clean$cpi_value[cpi_clean$report_year == 2022]

# 2. Join the tables and calculate "Real Income"
# We multiply 2010 wages by (2022_CPI / 2010_CPI) to see what they're worth today
final_income_data <- full_income_comparison %>%
  left_join(cpi_clean, by = "report_year") %>%
  mutate(real_income = ifelse(
    report_year == 2010,
    estimate * (cpi_2022 / cpi_2010),
    estimate
  ))

# 3. Take a look at the "Fair Comparison"
View(final_income_data)
```


If you look at your new final_income_data table, you'll see a column called real_income.
- For the 2022 rows, this is just their normal salary.
- For the 2010 rows, the number has been "inflated" to show what that money would be worth in 2022.

This tells us: "In 2010, a young person in Alabama made $21,086. But in today's money (inflation-adjusted), that felt like having $28,294."


```{r}
# This creates a much simpler table for your eyes
comparison_view <- final_income_data %>%
  select(NAME, variable, report_year, real_income) %>%
  pivot_wider(names_from = report_year, values_from = real_income) %>%
  mutate(diff = `2022` - `2010`)

View(comparison_view)
```


# Web Scraping
We are going to "scrape" the MIT Living Wage Calculator. This tells us the minimum amount of money a person needs to survive in each state.

Before we try to scrape all 50 states, let's make sure we can get the data from just one. We’ll look for the "Required Annual Income before Taxes" for a single adult with no children.

Scrape Alabama

```{r}
library(rvest)

# The URL for Alabama (FIPS code 01)
url <- "https://livingwage.mit.edu/states/01"

# Now R will know what read_html means!
webpage <- read_html(url)

# 2. Extract all tables from the page
# The 'Annual Expenses' table is usually the second one
tables <- webpage %>% html_table()
expenses_table <- tables[[2]]

# 3. Look at the data
print(expenses_table)
```

Scrape All States (The Loop)

(Since our project requires 50 states, we don't want to copy-paste that code 50 times. We'll use a loop to tell R: "Go to state 01, get the data, then go to state 02, and so on.")

```{r}
library(tidyverse)
library(rvest)
# 1. Get our list of unique state FIPS codes
fips_list <- unique(final_income_data$GEOID)

# 2. Create an empty list to store our results
all_living_wages <- list()

# 3. Start the loop (This might take a minute!)
# 1. Reset
all_living_wages <- list()
options(timeout = 300)

for (fips in fips_list) {
  state_url <- paste0("https://livingwage.mit.edu/states/", fips)
  
  try({
    page <- read_html(state_url)
    # Get EVERY table on the page
    tbls <- page %>% html_table(fill = TRUE)
    
    found_in_state <- FALSE
    
    # 2. Check every table one by one
    for (df in tbls) {
      # Search the first column (using [[1]] so the name doesn't matter)
      row_idx <- which(str_detect(df[[1]], regex("Required", ignore_case = TRUE)))
      
      if (length(row_idx) > 0) {
        # Found it! Extract and clean
        raw_text <- df[row_idx[1], 2]
        clean_num <- as.numeric(str_replace_all(as.character(raw_text), "[^0-9.]", ""))
        
        # Save to our list
        all_living_wages[[fips]] <- data.frame(GEOID = fips, living_wage = clean_num)
        print(paste("State:", fips, "| Found Wage:", clean_num))
        
        found_in_state <- TRUE
        break # Found what we need, stop looking at other tables for this state
      }
    }
    
    if (!found_in_state) {
      print(paste("Skipped State:", fips, "- Row not found in any table"))
    }
    
    Sys.sleep(2) # Keep it slow to avoid being blocked
    
  }, silent = TRUE)
}

# Combine
living_wage_final <- bind_rows(all_living_wages)
View(living_wage_final)

```


Now we are going to combine your Income data (from the Census/FRED) with your Living Wage data (from MIT). This is the final step of the "Data Acquisition" part of your project.

```{r}
# 1. Join the tables together
master_dataset <- final_income_data %>%
  left_join(living_wage_final, by = "GEOID")

# 2. Calculate the "Affordability Gap"
# Formula: (Real Income) - (Living Wage)
# Positive = Surviving | Negative = Struggling
master_dataset <- master_dataset %>%
  mutate(affordability_gap = real_income - living_wage)

# 3. View the master file
View(master_dataset)
```

Look at Row 1 (Alabama, Under 25, 2010):
- Real Income: $28,294 (This is what their 2010 paycheck feels like today).
- Living Wage: $36,084 (This is what they actually need to survive today).
- The Reality: Even with their wages adjusted for inflation, they are still $7,790 short of a survival budget.



To see the answer to the whole project in one tiny table, run this. It shows exactly how the "Gap" has changed for young people in every state:

```{r}
summary_comparison <- master_dataset %>%
  select(NAME, variable, report_year, affordability_gap) %>%
  pivot_wider(names_from = report_year, values_from = affordability_gap) %>%
  mutate(change_in_affordability = `2022` - `2010`)

View(summary_comparison)
```

Look at Row 3 (Alaska, Under 25):
- In 2010, the average young worker had about $10,542 in "breathing room" after paying for basic survival.
- In 2022, that dropped to $8,272.

Even though wages might look higher on paper, that worker has $2,270 less to spend on things like savings, travel, or a house than they did 12 years ago.


# Validation Table

```{r}
validation_table <- master_dataset %>%
  summarize(
    total_rows = n(),
    missing_living_wage = sum(is.na(living_wage)),
    impossible_incomes = sum(real_income <= 0, na.rm = TRUE),
    duplicate_check = sum(duplicated(paste(GEOID, variable, report_year)))
  )
print(validation_table)
```


