---
title: "ISA401 Final Project Doc"
author: "Team 10"
date: "2026-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Source 1: Census ACS Income (25-44)

# Sydney
```{r}
library(tidycensus)
library(tidyverse)

# Load the list of all available variables for 2024
v24 <- load_variables(2024, "acs1", cache = TRUE)

# Use RStudio's View to search keywords like "age", "income", or "rent"
View(v24)


income_vars <- c(income_25_through_44 = "B19049_003")

income_2010 <- get_acs(geography = "state", variables = income_vars, year = 2010, survey = "acs1")
income_2010$year <- 2010

income_2024 <- get_acs(geography = "state", variables = income_vars, year = 2024, survey = "acs1")
income_2024$year <- 2024

income_combined <- rbind(income_2010, income_2024)
income_final <- pivot_wider(income_combined, id_cols = NAME, names_from = year, names_prefix = "income_", values_from = estimate)
```


## Source 2: Census ACS Rent Data (25-44)

# Sydney
```{r}
rent_vars = c(median_rent ="B25064_001")
rent_2010 <- get_acs(geography = "state", variables = rent_vars, year = 2010, survey = "acs1")
rent_2010$year <- 2010
rent_2024 <- get_acs(geography = "state", variables = rent_vars, year = 2024, survey = "acs1")
rent_2024$year <- 2024

rent_combined <- rbind(rent_2010, rent_2024)

rent_final <- pivot_wider(rent_combined, id_cols = NAME, names_from = year, names_prefix = "rent_", values_from = estimate)
```


## Source 3: FRED CPI Data

# Annie
```{r}
library(tidyverse)
library(tidyquant)

# Pull CPI data from FRED
# CPIAUCSL = Consumer Price Index for All Urban Consumers
cpi_data <- tq_get("CPIAUCSL",
                   get = "economic.data",
                   from = "2010-01-01",
                   to = "2024-12-31")

# Add a year column
cpi_data$year <- format(cpi_data$date, "%Y")

# Calculate average CPI for 2010
cpi_2010 <- cpi_data %>%
  filter(year == "2010") %>%
  summarize(cpi = mean(price, na.rm = TRUE))

cpi_2010$year <- 2010

# Calculate average CPI for 2024
cpi_2024 <- cpi_data %>%
  filter(year == "2024") %>%
  summarize(cpi = mean(price, na.rm = TRUE))

cpi_2024$year <- 2024

# Combine 2010 and 2024 CPI data
cpi_combined <- rbind(cpi_2010, cpi_2024)

# Reshape data so 2010 and 2024 CPI values are side by side
cpi_final <- pivot_wider(cpi_combined,
                         names_from = year,
                         names_prefix = "cpi_",
                         values_from = cpi)

# Create inflation factor to adjust 2010 dollars into 2024 dollars
inflation_factor <- cpi_final$cpi_2024 / cpi_final$cpi_2010

cpi_final
inflation_factor
```


```{r}
ggplot(cpi_data, aes(x = date, y = price)) +
  geom_line() +
  labs(title = "Consumer Price Index from 2010 to 2024",
       x = "Year",
       y = "CPI")
```


The CPI trend shows that prices increased substantially between 2010 and 2024, with a sharp rise occurring after 2020. This supports the need to adjust earlier income and rent values for inflation before making affordability comparisons across years.

## Source 4: BLS OES Wage/Employment Data

# Cam


# Working on scraping all 50 states efficiently

```{r}
library(rvest)
library(httr2)
library(tidyverse)

# Check if allowed to scrape
robotstxt::paths_allowed(paths = "https://livingwage.mit.edu/states/01")

# All 50 state FIPS codes (zero-padded, skips territories/non-states)
state_fips <- c("01","02","04","05","06","08","09","10","12","13",
                "15","16","17","18","19","20","21","22","23","24",
                "25","26","27","28","29","30","31","32","33","34",
                "35","36","37","38","39","40","41","42","44","45",
                "46","47","48","49","50","51","53","54","55","56")

# Empty list to store results
all_states_tables <- list()

for (fips in state_fips) {
  
  url <- paste0("https://livingwage.mit.edu/states/", fips)
  
  page <- httr2::request(url) |>
    httr2::req_user_agent("An ISA 401 Student from Miami University") |>
    httr2::req_throttle(rate = 10/60) |>
    httr2::req_perform() |>
    httr2::resp_body_raw() |>
    rvest::read_html()
  
  tables <- page |>
    rvest::html_elements("table") |>
    rvest::html_table()
  
  tables <- lapply(tables, function(tbl) {
    tbl$state_fips <- fips
    tbl
  })
  
  all_states_tables[[fips]] <- tables
  
  message("Done: ", fips)
}

# Stack each table across all 50 states into flat data frames
living_wage_table1 <- map_dfr(all_states_tables, ~ .x[[1]])
living_wage_table2 <- map_dfr(all_states_tables, ~ .x[[2]])
living_wage_table3 <- map_dfr(all_states_tables, ~ .x[[3]])

# Add readable state names
state_xwalk <- tidycensus::fips_codes |>
  distinct(state_code, state_name) |>
  rename(state_fips = state_code)

living_wage_table1 <- living_wage_table1 |> left_join(state_xwalk, by = "state_fips")
living_wage_table2 <- living_wage_table2 |> left_join(state_xwalk, by = "state_fips")
living_wage_table3 <- living_wage_table3 |> left_join(state_xwalk, by = "state_fips")
```

## Merging All Sources

```{r}
# Step 1: Join income and rent by state name
merged_data <- income_final |>
  left_join(rent_final, by = "NAME")

# Step 2: Add CPI inflation factor as a constant column (it's national, not by state)
merged_data$inflation_factor <- inflation_factor

# Step 3: Add living wage data (table 1) by state name
# Rename state_name in the living wage crosswalk to match "NAME" used in Census data
living_wage_clean <- living_wage_table1 |>
  rename(NAME = state_name)

merged_data <- merged_data |>
  left_join(living_wage_clean, by = "NAME")

# Step 4: Adjust 2010 income and rent to 2024 dollars using inflation factor
merged_data$income_2010_adj <- merged_data$income_2010 * merged_data$inflation_factor
merged_data$rent_2010_adj   <- merged_data$rent_2010   * merged_data$inflation_factor

# Step 5: Calculate affordability ratios (annual rent / annual income)
# Rent is monthly, so multiply by 12 to annualize
merged_data$afford_ratio_2010 <- (merged_data$rent_2010_adj * 12)  / merged_data$income_2010_adj
merged_data$afford_ratio_2024 <- (merged_data$rent_2024 * 12)      / merged_data$income_2024

# Step 6: Calculate change in affordability (positive = worse, negative = better)
merged_data$afford_change <- merged_data$afford_ratio_2024 - merged_data$afford_ratio_2010

# View the final merged dataset
View(merged_data)
```

A few things to note:

**The affordability ratio** is annual rent divided by income. A common rule of thumb is that households should spend no more than 30% of income on rent — so a ratio above 0.30 signals a cost burden. `afford_change` tells you directly whether each state got worse or better between 2010 and 2024.

**`living_wage_table1`** needs to have a `state_name` column for the join to work, which it will since you added the crosswalk in Source 4. Just double-check the column names match by running `colnames(living_wage_table1)` and `colnames(income_final)` before running the merge.

**This final `merged_data` is what you export to Tableau/Power BI.** You can save it with:

```{r}
write.csv(merged_data, "final_merged_data.csv", row.names = FALSE)
```

